Examining equity in accessibility to bike share: a balanced floating catchment area approach

### 2021-08-30 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Proof reading Elise's edits before resubmitting R2; created document with marked changes between versions

  \title{Examining equity in accessibility to bike share: a balanced floating
catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) can play a role in advancing
  transportation equity if they make bicycling more accessible to
  disadvantaged populations. In Ontario, the City of Hamilton expanded
  their PBSP in 2018 by adding twelve ``equity'' stations with the
  explicit objective of increasing access for under-serviced
  neighbourhoods. In this case study, we investigate differentials in
  accessibility to stations using a balanced floating catchment area
  approach and compare accessibility with and without the equity stations.
  We analyze micro zones to better reflect walking to a station and
  conduct a sensitivity analysis at several walking time thresholds. We
  then reaggregate the estimated accessibility for further analysis using
  census data. Our findings indicate that equity stations increased
  accessibility for the serviced population at every threshold examined,
  but the increase was relatively modest especially for population in the
  bottom 20\% of median total household income.
  \end{abstract}

@@

The potential of public bicycle share programs (PBSPs) to increase
bicycling levels is but one of many reasons for implementing such
programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
inexpensive, and convenient form of public transportation, shared
bicycles can encourage individuals to take up bicycling for short local
trips or first and last mile trips to other public transportation
instead of using personal vehicles. These programs can also play a role
in advancing transportation equity if they make bicycling more
accessible to disadvantaged populations. Although PBSPs are available to
the general public in over 800 cities worldwide (Fishman, 2016), and
ought to be accessible to any individual who wishes to use them,
research on PBSPs indicates that inequities persist with respect to who
can use and access them.

Many PBSPs in North America now offer specific programs to address
equity that primarily focus on removing cost barriers and increasing
access for groups that are under-represented among existing users
(McNeil et al., 2019). The locations of docking stations is a common
consideration for increasing equity (Howland et al., 2017). Hamilton
Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
PBSP included in a North American scan of bike share equity programs
(see McNeil et al., 2019). HBS was launched in 2015 and currently has
over 900 operational bicycles and 130 docking stations. An equity
program, Everyone Rides Initiative, was implemented in 2018 which
expanded the program by introducing twelve ``equity'' stations to more
disadvantaged neighbourhoods in the core service area.

@@

This paper is an example of open and reproducible research that uses
only open software for transportation and statistical analysis (Bivand,
2020; Lovelace, 2021). All data were obtained from publicly available
sources and organized in the form of a data package. Following best
practices in spatial data science (Brunsdon and Comber, 2020), the code
and data needed to reproduce, modify or extend the analysis are
available for download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}

@@

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
However, one study found that bike share users in Washington, DC were
more likely to be female (Buck et al., 2013), which suggests that the
gender gap among bicyclists who use PBSPs is less disparate than the gap
for personal bicycle use (Fishman, 2016). There is some evidence that
bike share users are less likely to own a car (Buck et al., 2013;
Reilly, Noyes, et al., 2020). However, the relationship between income
or education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
Goodman, 2012). Being university educated was a significant correlate of
bike share use in Montreal, Canada (Fuller et al., 2011). Not
coincidentally, financial savings have been found to motivate those on a
low income to use bike share (Fishman, 2016).

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
(2020) estimated ridership in the city's PBSP and found that a minority
of bike share stations were located in disadvantaged communities, while
annual members from such areas had a lower share of trips compared to
other areas in the city. Similar results were found in Philadelphia.
Despite efforts to increase equity within the city's PBSP, census block
groups with lower median income generated fewer trips (Caspi and Noland,
2019). Trips from stations in such areas were utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al.~(2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al.~(2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where bike share stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location matters. Kabra et al.~(2020) found that the majority of bike
share usage in Paris, France comes from areas within 300 m of stations,
which amounts to 2-4 minutes walking by an adult who does not have a
disability. Similar to other public amenities affected by crowding, the
utility of docking stations is also limited by the \emph{maximum} number
of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bike share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). The gravity-based
approach involves weighting destination opportunities, such as the
quantity of bike share stations, by the time required to reach them from
an origin using an impedance function (Handy and Niemeier, 1997; Kwan,
1998). While such measures are suitable for capturing the potential for
reaching destinations from a given location, they do not take demand or
congestion effects into account which is an important consideration when
calculating accessibility for amenities such as bike share stations.

In contrast, floating catchment area (FCA) methods have been widely
employed in health care accessibility research. The benefit of this
approach is that the method incorporates information on capacity,
demand, and congestion in calculating accessibility. This approach is
more appropriate and informative than calculating provider-to-population
ratios (PPR) that simply divide the level of supply of a service (e.g.,
the number of bicycle racks at a station) by the population who have
access to the service (Paez et al., 2019). In particular, the Two-Step
Floating Catchment Area (2SFCA) method (Luo and Wang, 2003; Radke and
Mu, 2000) produces flexible catchment areas instead of using rigid
boundaries like PPR. In the first step, a ratio of supply to demand at
service locations is calculated, such as the number of beds at a
hospital divided by the number of people within the catchment area of
the hospital, weighted by the distance involved in reaching the
facility. Next, these service level ratios are allocated back to the
population centers and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. However, overlapping catchment
areas from the conventional FCA approach lead to inflation of population
totals and deflation of service levels across a study area which
generates inaccurate or misleading accessibility estimates (Paez et al.,
2019). While there have been many methodological innovations in FCA
methods (e.g., Delamater, 2013; Luo and Qi, 2009; Luo and Wang, 2003;
Radke and Mu, 2000; Wan et al., 2012), a recent improvement to this
approach was achieved through a simple and intuitive balancing of the
impedance that addressed the effects of demand and service inflation
found in earlier FCA approaches (see Paez et al., 2019). This provides
more useful information because it does not assume that people are
limited to service within pre-defined boundaries (Paez et al., 2019).

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

@@

Hamilton Bike Share Inc., the non-profit operator, launched an equity
program, Everyone Rides Initiative (ERI), in 2018 with the objective of
reducing barriers that may prevent individuals from accessing the
program. Additional bicycles and twelve ``equity'' stations were added
to the core service area in more disadvantaged neighbourhoods. The
equity program also offers subsidized memberships to individuals who
identify as low income, and complements this service with cycle skills
education. A comparable program can be found in Philadelphia (see Caspi
and Noland, 2019).

@@

The BFCA method was developed to address issues with demand and supply
inflation that result from the overlapping catchment areas produced by
earlier FCA methods (Paez et al., 2019). Paez et al.~(2019) adjusted the
impedance weights so that both supply and demand are proportionally
allocated. The result is a FCA method that balances the population and
level of service by eliminating the over-counting of population and
level of service that leads to distortions in demand and supply.

@@

As seen in the equation above, the population allocated to station \(j\)
is the weighted sum of the population in the region; a spatial weight
\(w_{ij}\) represents the friction that the population at \(i\) faces
when reaching station \(j\), and is usually given by a distance-decay
function, so that each station is assumed to service only a segment of
the population within a limited geographical range.

Next, the supply at each station (i.e., the maximum number of bicycle
racks) is divided by its estimated service population within the
established catchment area; this gives the level of service of station
\(j\) in bicycle racks per person: \begin{equation}
\label{eq:level-of-service}
L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\end{equation}

Finally, the accessibility of population cell \(i\) is calculated as the
weighted sum of the level of service of all stations that can be reached
from there according to the spatial weights: \begin{equation}
\label{eq:FCA-accessibility}
A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\end{equation}

The balanced approach of Paez et al (2019) replaces the spatial weights
with normalized versions as follows: \begin{equation}
\label{eq:spatial-weights}
{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
\end{equation}

\noindent and: \begin{equation}
\label{eq:spatial-weights-2}
{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
\end{equation}

These weights satisfy the following properties: \begin{equation}
\label{eq:weights}
\sum_{j = 1}^{J} {w^i_{ji}} = 1
\end{equation}

\noindent and: \begin{equation}
\label{eq:weights-2}
\sum_{i = 1}^{n} {w^j_{ji}} = 1
\end{equation}

@@

Since walking trips to a docking station likely happen at a level lower
than even the smallest census geography, this requires a more
disaggregated approach than the use of census geographies. For this
reason, we implemented our analysis parting from micro population zones
to better reflect the friction of walking to a docking station, which is
an important component of a bike share trip (Chen et al., 2019).

To obtain population at sub-census geography levels, we used
pycnophylactic interpolation (Tobler, 1979). We obtained population data
from the 2016 Census of Canada for dissemination areas (DA), which is
the smallest publicly available census geography in Canada. These zonal
values of the population were interpolated to smaller polygons that are
50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
the population from each DA while preserving total volume {[}see Figure
\ref{fig:da-population}{]}. When interpolating the population at this
high level of resolution, it is important to ensure that population
numbers were not allocated to areas where people do not live in Hamilton
(for example, to parks or large institutional buildings, etc.). To do
so, we retrieved shapefiles for various geographic features from Open
Hamilton. Next, we removed these features from the PBSP core service
area and used pycnophylactic interpolation to disaggregate and
reallocate population within the remaining area {[}see Figure
\ref{fig:interpolated-population}{]}.

@@

To calculate walking times from the centroid of the micro population
cells to docking stations, we extracted OpenStreetMap data for HBS's
service area from
\href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
route planner that interfaces with OpenStreetMap. OpenStreetMap data
provides the networks for calculating walking times from each population
cell to nearby docking stations, using a maximum walking distance of 10
km and walking time of 30 minutes as thresholds. A travel time matrix
was created with the origins as the coordinates of the population cells
and the destinations as the coordinates of the docking stations within
the maximum threshold. This process provides a more realistic measure of
the friction of reaching stations by taking infrastructure into account
network travel times, rather than using the Euclidean distance from
population cell to station. Routing and travel time calculations were
completed using the \texttt{R} package \texttt{r5r}, used for rapid
realistic routing operations (Pereira et al., 2021).

@@

Consensus regarding the distance that individuals are willing to walk to
access a docking station is lacking, but the literature on walking
behaviour provided some guidelines to determine the thresholds for our
sensitivity analysis. Previous studies have found that living within 250
m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
station is correlated with bike share use, while other research has
found that walking trips are less than 600 m and rarely more than 1200 m
(Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
2010). HBS will often depict a map at some docking stations to show the
locations of the other nearest stations within a five minute walk, which
suggests that this is an average distance that people are expected to
walk. The National Association of City Transportation Officials (NACTO)
has a similar normative guide (City Transportation Officials, 2015).

@@

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share, with a particular focus on assessing the
contribution of the program's equity stations. The BFCA approach,
combined with pycnophylactic interpolation, enabled us to measure
accessibility on a micro scale which better reflects the scale at which
walking takes place and avoids the ``absorption of disparities'', as
articulated by Chen et al.~(2019) and elsewhere in the transportation
literature (National Academies of Sciences, 2004; Rowangould et al.,
2016). Our method also considers demand and supply in the calculation of
accessibility, which is important for services like PBSPs but not
accounted for in other common accessibility models. Unlike the 2SFCA
method, there was no inflation or deflation in our accessibility
estimates because population and level of service were preserved (see
Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
differentiates our analysis from similar papers exploring equity in
PBSPs that use larger geographical units of analysis or that focus on
station location instead of level of service. In this way, our paper has
made an important contribution by applying an intuitive and useful
approach to measure accessibility to a PBSP. This open and reproducible
research could be expanded for additional analysis using other
individual-level data in Hamilton (e.g., age, gender, household size,
education, etc.) or applied to PBSPs in other North American cities.

>>>>

  \title{Examining equity in accessibility to bike share: a balanced
floating catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Páez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \&
Society, McMaster University, 1280 Main Street West, Hamilton, ON
L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography
\& Planning, University of Toronto Scarborough, 1265 Military Trail,
Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) can play a role in advancing
  transportation equity if they make bicycling more accessible to
  disadvantaged populations. In Ontario, Hamilton Bike Share expanded
  their program in 2018 by adding twelve ``equity'' stations with the
  explicit objective of increasing access for under-serviced
  neighbourhoods. In this case study, we investigate differentials in
  accessibility to stations using a balanced floating catchment area
  approach and compare accessibility with and without the equity
  stations. We analyze population interpolated to small cells to better
  reflect walking to a station and conduct a sensitivity analysis at
  several walking time thresholds. We then reaggregate the estimated
  accessibility by income groups for further analysis. Our findings
  indicate that equity stations increased accessibility for the serviced
  population at every threshold examined, but the increase was
  relatively modest especially for population in the bottom 20\% of
  median total household income.
  \end{abstract}

@@

The potential of public bicycle share programs (PBSPs) to increase
bicycling levels is but one of many reasons for implementing such
programs in urban areas (see Fishman et al., 2015; Hosford et al., 2019,
2018). As a healthy, inexpensive, and convenient form of public
transportation, shared bicycles can encourage individuals to take up
bicycling for short local trips or first and last mile trips to other
public transportation instead of using personal vehicles. These programs
can also play a role in advancing transportation equity if they make
bicycling more accessible to disadvantaged populations. Although PBSPs
are available to the general public in over 800 cities worldwide
(Fishman, 2016), and ought to be accessible to any individual who wishes
to use them, research on PBSPs indicates that inequities persist with
respect to who can use and access them.

Many PBSPs in North America now offer specific programs to address
equity that primarily focus on removing cost barriers and increasing
access for groups that are under-represented among existing users
(McNeil et al., 2019). The locations of docking stations is a common
consideration for increasing access to improve equity (Howland et al.,
2017). Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the
only Canadian PBSP included in a North American scan of bike share
equity programs (see McNeil et al., 2019). HBS was launched in 2015 and
currently has over 900 operational bicycles and 130 docking stations. An
equity program, \emph{Everyone Rides Initiative}, was implemented in
2018 which expanded the program by introducing twelve ``equity''
stations to more disadvantaged neighbourhoods in the core service area.

@@

This paper is an example of open and reproducible research that uses
only open software for transportation and statistical analysis (Bivand,
2020; Lovelace, 2021). All data were obtained from publicly available
sources and organized in the form of a data package. Following best
practices in spatial data science (Brunsdon and Comber, 2020), an open
data product (Arribas/Bel et al., 2021) along with the code needed to
reproduce, modify or extend the analysis are available for
download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}

@@

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users to
date (Fishman, 2016). In many cities, males use bike share more than
females (Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman,
2012; Reilly et al., 2020b; Wang and Akar, 2019; Winters et al., 2019)
as do younger age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller
et al., 2011). However, one study found that bike share users in
Washington, DC were more likely to be female (Buck et al., 2013), which
suggests that the gender gap among bicyclists who use PBSPs is less
disparate than the gap for personal bicycle use (Fishman, 2016). There
is some evidence that bike share users are less likely to own a car
(Buck et al., 2013; Reilly et al., 2020a). However, the relationship
between income or education and bike share use is less clear-cut.
Stations in disadvantaged communities in Chicago have been found to
generate most of the average annual trips (Qian and Jaller, 2020) and
individuals from minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul also used the city's PBSP more (Wang and Lindsey,
2019a). Similar findings were reported in London (Ogilvie and Goodman,
2012). Being university educated was a significant correlate of bike
share use in Montreal, Canada (Fuller et al., 2011). Not coincidentally,
financial savings have been found to motivate those on a low income to
use bike share (Fishman, 2016).

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on who benefits from them. Many studies have examined
differences in demographics or socioeconomic status between those who
use or have access to PBSPs and those who don't, while other research
has explored spatial inequities in where stations are located (e.g.,
Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et al.,
2019; Qian and Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian
and Jaller (2020) estimated ridership in the city's PBSP and found that
a minority of docking stations were located in disadvantaged
communities, while annual members from such areas had a lower share of
trips compared to other areas in the city. Similar results were found in
Philadelphia. Despite efforts to increase equity within the city's PBSP,
census block groups with lower median income generated fewer trips
(Caspi and Noland, 2019). Trips from stations in such areas were
utilitarian (e.g., commuting to work), which points to the importance of
ensuring equitable access (Caspi and Noland, 2019). In the case of
Seattle, all neighbourhoods were found to have some level of access to
dockless bicycles but those with higher incomes and more residents of
higher education had more bicycles (Mooney et al., 2019). Babagoli et
al. (2019) also found that neighbourhoods in New York City with higher
affluence had the greatest proportion of Citi Bike stations.

Chen et al. (2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (Chen et al., 2019). Both
are of interest to researchers and transportation planners since they
are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs (Howland et al., 2017; see
McNeil et al., 2019) can be considered efforts to improve vertical
equity since they favour groups that have benefited less from PBSPs
through the placement of stations or equitable fee structures.

On the whole, existing studies highlight the need for PBSPs to be more
accessible for equity populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where docking stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location matters. Kabra et al. (2020) found that the majority of bike
share usage in Paris, France comes from areas within 300 m of stations,
which amounts to 2-4 minutes walking by an adult who does not have a
disability. Other studies have adopted (2019, 2018) Similar to other
public amenities affected by crowding, the utility of docking stations
is also limited by the \emph{maximum} number of bicycles that they can
hold (e.g., their size). Accessibility analyses for PBSPs constitute a
positive and evaluation-based approach that also has the potential to
inform equity efforts. For instance, Wang and Lindsey (2019b)
investigated whether new or relocated bike share stations increased
accessibility and use, which offered important insights to improve the
performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Geurs and van Wee, 2004; Handy and Niemeier,
1997). The gravity-based approach involves weighting destination
opportunities, such as the quantity of bike share stations, by the time
required to reach them from an origin using an impedance function (Handy
and Niemeier, 1997; Kwan, 1998). While such measures are suitable for
capturing the potential for reaching destinations from a given location,
they do not take demand or congestion effects into account which is an
important consideration when calculating accessibility for amenities
such as bike share stations.

In contrast, floating catchment area (FCA) methods have been widely
employed in health care accessibility research. The benefit of this
approach is that the method incorporates information on capacity and
demand in calculating accessibility. Geurs and van Wee (2004) note
``competition for activities with restricted capacity'' should be taken
into account for land-use components of accessibility. The FCA approach
is more appropriate and informative than calculating
provider-to-population ratios (PPR) that simply divide the level of
supply of a service (e.g., the number of bicycle racks at a station) by
the population who have access to the service (Paez et al., 2019). In
particular, the Two-Step Floating Catchment Area (2SFCA) method (Luo and
Wang, 2003; Radke and Mu, 2000) produces flexible catchment areas
instead of using rigid boundaries like PPR. In the first step, a ratio
of supply to demand at service locations is calculated, such as the
number of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centers and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. However, overlapping catchment
areas from the conventional FCA approach lead to inflation of population
totals and deflation of service levels across a study area which
generates inaccurate or misleading accessibility estimates (Paez et al.,
2019). While there have been many methodological innovations in FCA
methods (e.g., Delamater, 2013; Luo and Qi, 2009; Luo and Wang, 2003;
Radke and Mu, 2000; Wan et al., 2012), a recent improvement to this
approach was achieved through a simple and intuitive balancing of the
impedance that addressed the effects of demand and service inflation
found in earlier FCA approaches (see Paez et al., 2019). This provides
more useful information because it does not assume that people are
limited to service within pre-defined boundaries (Paez et al., 2019).

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al. (2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation.''

@@

Hamilton Bike Share Inc., the non-profit operator, launched an equity
program, Everyone Rides Initiative (ERI), in 2018 with the objective of
reducing barriers that may prevent individuals from accessing the
program. Additional bicycles and twelve ``equity'' stations were added
to the core service area in more disadvantaged and under-serviced
neighbourhoods. The equity program also offers subsidized memberships to
individuals who identify as low income, and complements this service
with cycle skills education. A comparable program can be found in
Philadelphia (see Caspi and Noland, 2019).

@@

The BFCA method was developed to address issues with demand and supply
inflation that result from the overlapping catchment areas produced by
earlier FCA methods (Paez et al., 2019). Páez et al. (2019) adjusted the
impedance weights so that both supply and demand are proportionally
allocated. The result is a FCA method that balances the population and
level of service by eliminating the over-counting of population and
level of service that leads to distortions in demand and supply.

@@

As seen in the equation above, the population allocated to station \(j\)
is the weighted sum of the population in the region; a spatial weight
\(w_{ij}\) represents the friction that the population at \(i\) faces
when reaching station \(j\), and is usually given by a distance-decay
function, so that each station is assumed to service only a segment of
the population within a limited geographical range. The level of service
of station \(j\) in bicycle racks per person is the supply at each
station (i.e., the maximum number of bicycle racks) divided by the
population within the established catchment area: \begin{equation}
\label{eq:level-of-service}
L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\end{equation}

In the second step, the accessibility of population unit \(i\) is
calculated as the weighted sum of the level of service of all stations
that can be reached from there according to the spatial weights:
\begin{equation}
\label{eq:FCA-accessibility}
A_i = {\sum_{j = 1}^{J} L_j{w_{ij}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ij}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\end{equation}

The balanced approach of Páez et al. (2019) replaces the spatial weights
with normalized versions as follows. In the first step, the population
is weighted with: \begin{equation}
\label{eq:spatial-weights}
{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}
\end{equation}

\noindent and in the second step, the level of service is weighted
using: \begin{equation}
\label{eq:spatial-weights-2}
{w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}
\end{equation}

These weights satisfy the following properties: \begin{equation}
\label{eq:weights}
\sum_{j = 1}^{J} {w^i_{ij}} = 1
\end{equation}

\noindent and: \begin{equation}
\label{eq:weights-2}
\sum_{i = 1}^{n} {w^j_{ij}} = 1
\end{equation}

@@

Since walking trips to a docking station likely happen at a level lower
than even the smallest census geography, this requires a more
disaggregated approach than the use of census geographies. For this
reason, we implemented our analysis parting from small population cells
to better reflect the friction of walking to a docking station, which is
an important component of a bike share trip (Chen et al., 2019).

To obtain population at sub-census geography levels, we used
pycnophylactic interpolation (Tobler, 1979). We obtained population data
from the 2016 Census of Canada for dissemination areas (DA), which is
the smallest publicly available census geography in Canada. These zonal
values of the population were interpolated to smaller polygons that are
50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
the population from each DA while preserving total volume. When
interpolating the population at this high level of resolution, it is
important to ensure that population numbers are not allocated to areas
where people do not live in Hamilton (for example, to parks or large
institutional buildings, etc.). To do so, we retrieved shapefiles for
various geographic features from Open Hamilton. Next, we removed these
features from the PBSP core service area and used pycnophylactic
interpolation to disaggregate and reallocate population within the
remaining area {[}see Figure \ref{fig:interpolated-population}{]}.

@@

To calculate walking times from the centroid of the population cells to
docking stations, we extracted OpenStreetMap data for HBS's service area
from \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online
cycle route planner that interfaces with OpenStreetMap. OpenStreetMap
data provides the networks for calculating walking times from each
population cell to nearby docking stations, using a maximum walking
distance of 10 km and walking time of 30 minutes as thresholds. A travel
time matrix was created with the origins as the coordinates of the
population cells and the destinations as the coordinates of the docking
stations within the maximum threshold. This process provides a more
realistic measure of the friction of reaching stations by taking
infrastructure into account network travel times, rather than using the
Euclidean distance from population cell to station. Routing and travel
time calculations were completed using the \texttt{R} package
\texttt{r5r}, used for rapid realistic routing operations (Pereira et
al., 2021).

@@

Consensus regarding the distance that individuals are willing to walk to
access a docking station is lacking, but the literature on walking
behaviour provided some guidelines to determine the thresholds for our
sensitivity analysis. Previous studies have found that living within 250
m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
station is correlated with bike share use. Some studies have defined
``close proximity'' to a docking station as 500 metres or less (Fuller
et al., 2013; Hosford et al., 2019, 2018), while other research has
found that walking trips are less than 600 m and rarely more than 1200 m
(Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
2010). HBS will often depict a map at some docking stations to show the
locations of the other nearest stations within a five minute walk, which
suggests that this is an average distance that people are expected to
walk. The National Association of City Transportation Officials (NACTO)
has a similar normative guide (City Transportation Officials, 2015).

@@

In the analysis we use the following distance decay function: \[
{w_{ij} = \begin{cases} 
      1 & \text{when } t_{ij}\leq \gamma \\
      0 & \text{otherwise} 
   \end{cases}}
\] \noindent where \(\gamma\) is the relevant threshold. The weights are
standardized as discussed above.

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share (HBS), with a focus on assessing the contribution
of the program's equity stations. The BFCA approach, combined with
pycnophylactic interpolation, enabled us to measure accessibility on a
micro scale which better reflects the scale at which walking takes place
and avoids the ``absorption of disparities,'' as articulated by Chen et
al. (2019) and elsewhere in the transportation literature (National
Academies of Sciences, 2004; Rowangould et al., 2016). Our method also
considers potential demand and supply, as well as congestion effects, in
the calculation of accessibility, which is important for services like
PBSPs but not accounted for in other common accessibility models. Unlike
the 2SFCA method, there was no inflation or deflation in our
accessibility estimates because population and level of service were
preserved (see Figure \ref{fig:figure-10} and Figure
\ref{fig:figure-11}). This differentiates our analysis from similar
papers exploring equity in PBSPs that use larger geographical units of
analysis or that focus only on station location instead of level of
service. In this way, our paper has made an important contribution by
applying an intuitive and useful approach not previously used in the
cycling literature to measure accessibility to a PBSP. This open and
reproducible research could be expanded for additional analysis using
other individual-level data in Hamilton (e.g., age, gender, household
size, education, etc.) or applied to PBSPs in other North American
cities.
